---
title: "From three disagreeing sources to one patient-count estimate"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{From three disagreeing sources to one patient-count estimate}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", fig.width = 6.5, fig.height = 4
)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
```

## The problem this package is for

A recruiter, a client, or a health ministry rarely wants a rate. They want a
number of people: how many patients have this condition, in this region,
this year, and in five years. Getting there from public data usually means
reconciling two or three sources that do not quite agree -- a household
survey, an administrative registry, a model-based estimate -- and being
honest about how much that disagreement should widen the final interval.

This vignette walks the whole pipeline on a **synthetic** dataset shaped
like that problem: six fictional regions, an age-stratified population for
each, and three independent prevalence estimates per region for a fictional
chronic condition. None of the numbers below describe a real disease,
country, or published estimate -- swap `simulate_burden_example()` for real
WHO/IHME/national-survey exports that share the same column names and
everything downstream works unchanged.

```{r setup}
library(popburden)

d <- simulate_burden_example()
head(d$population)
d$sources
```

## Step 1: reconcile the sources, region by region

Each region has three prevalence estimates that disagree by design --
`administrative_registry` runs low (registries undercount), the others
cluster closer to the truth. `reconcile_estimates()` pools them with
inverse-variance weights and, by default, lets the between-source spread
inflate the pooled interval (DerSimonian-Laird random effects) rather than
pretending three different measurement processes share one error model.

```{r}
regions <- unique(d$sources$region)

reconciled <- lapply(regions, function(r) {
  s <- d$sources[d$sources$region == r, ]
  reconcile_estimates(s$estimate, s$se, source = s$source)
})
names(reconciled) <- regions

reconciled[["Region A"]]
```

```{r, eval = has_ggplot2, fig.alt = "Forest plot of three prevalence sources and their pooled estimate for Region A"}
plot_reconciled(reconciled[["Region A"]])
```

## Step 2: turn the pooled rate into a patient count

`estimate_population_count()` propagates the pooled rate's uncertainty
through to an absolute count via Monte Carlo simulation, sampling from a
Beta distribution so the simulated rate never leaves \[0, 1\].

```{r}
region_totals <- stats::aggregate(population ~ region, d$population, sum)

pop_table <- do.call(rbind, lapply(regions, function(r) {
  pooled <- reconciled[[r]]
  pop <- region_totals$population[region_totals$region == r]
  out <- estimate_population_count(
    rate = pooled$estimate, rate_se = pooled$se,
    population = pop, distribution = "beta"
  )
  out$region <- r
  out
}))
pop_table[, c("region", "population", "rate", "sim_mean", "ci_low", "ci_high")]
```

That table is the deliverable: six regions, one reconciled prevalence and
one patient-count estimate with an interval per region, instead of eighteen
disagreeing numbers.

## Step 3: age-standardize when regions differ in age structure

Raw prevalence comparisons across regions can just reflect who is old,
not how sick a region is. `age_standardize()` computes a directly
standardized rate against whatever reference-population weights you
supply, with two ways to get a confidence interval: a fast normal
approximation, or a Poisson parametric bootstrap for when some strata have
only a handful of events.

```{r}
strata <- data.frame(
  age_band   = c("0-14", "15-44", "45-64", "65+"),
  count      = c(3, 14, 52, 96),
  population = c(180000, 260000, 110000, 54000)
)
# Illustrative equal weights; use a published reference population
# (e.g. the WHO World Standard Population) for a real analysis.
age_standardize(
  strata$count, strata$population,
  weight = c(0.25, 0.25, 0.25, 0.25), method = "bootstrap"
)
```

## Step 4: project forward

`project_forward()` is deliberately simple: it compounds a population
growth assumption and a rate-trend assumption forward from a base-year
count, with both assumptions allowed their own uncertainty. It is a
starting point for a scenario conversation, not a replacement for a full
demographic or claims-based forecast.

```{r}
national_base <- sum(pop_table$sim_mean)

proj <- project_forward(
  base_count = national_base, horizon_years = 5,
  pop_growth = 0.008, pop_growth_sd = 0.002,
  rate_trend = 0.015, rate_trend_sd = 0.01
)
proj
```

```{r, eval = has_ggplot2, fig.alt = "Fan chart of the projected national patient count over five years"}
plot_projection(proj, base_year = 2026)
```

## Summary

- `reconcile_estimates()` -- pool disagreeing source estimates, with
  between-source heterogeneity reflected in the interval.
- `age_standardize()` -- compare regions or years on a level age basis.
- `estimate_population_count()` -- turn a rate into a patient count without
  losing the uncertainty.
- `project_forward()` -- extend a current estimate into a multi-year
  scenario.

Every function returns both a point estimate and an interval; none of them
silently drop the uncertainty on the way to a headline number.
